---
title: "Post-Hurricane  Vital  Statistics  Expose  Fragility  of  Puerto  Rico's  Health  System"
subtitle: Computations for the Paper
output:
  html_document:
    df_print: paged
  pdf_document: default
---
Let's start by loading the data:

```{r}
source("init.R")
load("rdas/fit-models.rda")
```

### Abstract: Results

Here are percent changes in mortality for each hurricane:

```{r}
combined_daily_res %>% ### Hurricanes in PR
  select(Date, f_hat, state) %>% 
  mutate(percent_change = 100*(exp(f_hat)-1)) %>%
  filter(Date == hurricane_dates$PR[1] |
         Date == hurricane_dates$PR[2] |
         Date == hurricane_dates$PR[3] &
         state == "PR")

combined_daily_res %>% ### Hurricane Irma
  select(Date, f_hat, state) %>% 
  mutate(percent_change = 100*(exp(f_hat)-1)) %>%
  filter(Date == hurricane_dates$FL &
         state == "FL")

combined_monthly_res %>% ### Hurricane Katrina
  select(Date, increase, state) %>% 
  mutate(percent_change = 100*increase) %>%
  filter(Date == hurricane_dates$LA &
         state == "LA")
```


Here are the prolonged increase in death rate after Maria and katrina, respectively:

```{r}
maria <- combined_daily_res %>% # Hurricane Maria
          filter(state == "PR" & Date >= hurricane_dates$PR[3] & f_hat - 1.96*se <= 0) %>% 
          .$Date %>% first() %>% as.numeric(.) - as.numeric(hurricane_dates$PR[3])
maria

georges <- combined_daily_res %>% # Hurricane Georges
            filter(state == "PR" & Date >= hurricane_dates$PR[2] & f_hat - 1.96*se <= 0) %>% 
            .$Date %>% first() %>% as.numeric(.) - as.numeric(hurricane_dates$PR[2])
georges

katrina <- combined_monthly_res %>% # Hurricane Katrina
            filter(state == "LA" & 
                     Date >= hurricane_dates$LA & 
                     increase - 1.96*se/(s_hat+year_offset) <= 0) %>% 
            .$Date %>% first() %>% as.numeric(.) - as.numeric(hurricane_dates$LA)
katrina
```
 
and here are the the point estimates and 95% CI for each of them:

```{r}
combined_daily_res %>% # Hurricane Maria
    filter(state == "PR" & graph_year == year(hurricane_dates$PR[3]) &
             Date >= hurricane_dates$PR[3]) %>%
    mutate(excess = cumsum(Deaths - exp(offset_sample_size+offset_seasonal+offset_year)),
           days = t - min(t) + 1,
           se = sqrt(exp(offset_sample_size+offset_seasonal+offset_year)),
           excess_se=sqrt(cumsum(se^2)),
           lower_CI = excess-1.96*excess_se,
           upper_CI = excess+1.96*excess_se) %>%
    select(days, excess, excess_se, lower_CI, upper_CI, hurricane) %>%
    bind_rows(tibble(days=0, excess=0, excess_se=0, hurricane = hurricane_names$PR[3])) %>%
    arrange(days) %>% filter(days==maria)

combined_daily_res %>% # Hurricane Georges
    filter(state == "PR" & graph_year == year(hurricane_dates$PR[2]) &
             Date >= hurricane_dates$PR[2]) %>%
    mutate(excess = cumsum(Deaths - exp(offset_sample_size+offset_seasonal+offset_year)),
           days = t - min(t) + 1,
           se = sqrt(exp(offset_sample_size+offset_seasonal+offset_year)),
           excess_se=sqrt(cumsum(se^2)),
           lower_CI = excess-1.96*excess_se,
           upper_CI = excess+1.96*excess_se) %>%
    select(days, excess, excess_se, lower_CI, upper_CI, hurricane) %>%
    bind_rows(tibble(days=0, excess=0, excess_se=0, hurricane = hurricane_names$PR[3])) %>%
    arrange(days) %>% filter(days==georges)


combined_monthly_res %>% # Hurricane Katrina
              filter(state == "LA" & graph_year == year(hurricane_dates$LA[1]) &
                     Date >= hurricane_dates$LA[1]) %>%
              mutate(excess = cumsum(Deaths - (s_hat + year_offset)*Population*days/365),
                     se = sqrt((s_hat + year_offset)*Population*days/365),
                     excess_se=sqrt(cumsum(se^2)),
                     lower_CI = excess-1.96*excess_se,
                     upper_CI = excess+1.96*excess_se,
                     days = cumsum(days)) %>%
              select(days, excess, excess_se, lower_CI, upper_CI, hurricane) %>%
              bind_rows(tibble(days=0, excess=0, hurricane = "LA: Katrina")) %>%
              arrange(days) %>% filter(days==katrina)
```

### Results: Indirect effects

Here is the death rate increase after Hurricane Maria with 95% CI
```{r}
combined_daily_res %>% ### Hurricanes in PR
  select(Date, f_hat, se, state) %>% 
  mutate(percent_change = 100*(exp(f_hat)-1)) %>%
  filter(Date == hurricane_dates$PR[3] &
         state == "PR") %>%
  mutate(lower_CI = 100*(exp(f_hat-1.96*se)-1),
         upper_CI = 100*(exp(f_hat+1.96*se)-1))
```

When did the death rate returned to historical levels after maria:

```{r}
combined_daily_res %>% # Hurricane Maria
          filter(state == "PR" & Date >= hurricane_dates$PR[3] & f_hat - 1.96*se <= 0) %>% 
          .$Date %>% first()
```

Average increase from Sep 20, 2007 to April 15, 2008:

```{r}
combined_daily_res %>% ### Hurricanes in PR
  select(Date, f_hat, se, state) %>% 
  mutate(percent_change = 100*(exp(f_hat)-1)) %>%
  filter(state == "PR" & Date >= hurricane_dates$PR[3] & f_hat - 1.96*se >= 0) %>%
  .$percent_change %>%
  mean(.)
```

Deaths the day that katrina made landfall and % increase:

```{r}
combined_monthly_res %>% 
              filter(state == "LA" & Date == hurricane_dates$LA) %>%
  mutate(percent_change = 100*increase) %>%
  select(Deaths, percent_change)
```

% change in mortality on the four months after Katrina:

```{r}
combined_monthly_res %>% 
              filter(state == "LA" & Date >= hurricane_dates$LA)%>%
    mutate(percent_change = 100*increase) %>%
    select(Date, Deaths, percent_change)
```

Here is the death rate increase after Hurricane Georges with 95% CI

```{r}
combined_daily_res %>% ### Hurricanes in PR
  select(Date, f_hat, se, state) %>% 
  mutate(percent_change = 100*(exp(f_hat)-1)) %>%
  filter(Date == hurricane_dates$PR[2] &
         state == "PR") %>%
  mutate(lower_CI = 100*(exp(f_hat-1.96*se)-1),
         upper_CI = 100*(exp(f_hat+1.96*se)-1))
```

When did the death rate returned to historical levels after georges:

```{r}
combined_daily_res %>% # Hurricane Maria
          filter(state == "PR" & Date >= hurricane_dates$PR[2] & f_hat - 1.96*se <= 0) %>% 
          .$Date %>% first()
```

Average increase from Georges

```{r}
combined_daily_res %>% ### Hurricanes in PR
  select(Date, f_hat, se, state) %>% 
  mutate(percent_change = 100*(exp(f_hat)-1)) %>%
  filter(state == "PR" & Date >= hurricane_dates$PR[2] & f_hat - 1.96*se >= 0) %>%
  .$percent_change %>%
  mean(.)
```

### Results: Excess deaths

Estimates of excess deaths

```{r}
combined_daily_res %>% # Hurricane Maria
    filter(state == "PR" & graph_year == year(hurricane_dates$PR[3]) &
             Date >= hurricane_dates$PR[3]) %>%
    mutate(excess = cumsum(Deaths - exp(offset_sample_size+offset_seasonal+offset_year)),
           days = t - min(t) + 1,
           se = sqrt(exp(offset_sample_size+offset_seasonal+offset_year)),
           excess_se=sqrt(cumsum(se^2)),
           lower_CI = excess-1.96*excess_se,
           upper_CI = excess+1.96*excess_se) %>%
    select(days, excess, excess_se, lower_CI, upper_CI, hurricane) %>%
    bind_rows(tibble(days=0, excess=0, excess_se=0, hurricane = hurricane_names$PR[3])) %>%
    arrange(days) %>% filter(days==maria)

combined_daily_res %>% # Hurricane Georges
    filter(state == "PR" & graph_year == year(hurricane_dates$PR[2]) &
             Date >= hurricane_dates$PR[2]) %>%
    mutate(excess = cumsum(Deaths - exp(offset_sample_size+offset_seasonal+offset_year)),
           days = t - min(t) + 1,
           se = sqrt(exp(offset_sample_size+offset_seasonal+offset_year)),
           excess_se=sqrt(cumsum(se^2)),
           lower_CI = excess-1.96*excess_se,
           upper_CI = excess+1.96*excess_se) %>%
    select(days, excess, excess_se, lower_CI, upper_CI, hurricane) %>%
    bind_rows(tibble(days=0, excess=0, excess_se=0, hurricane = hurricane_names$PR[3])) %>%
    arrange(days) %>% filter(days==georges)


combined_monthly_res %>% 
              filter(state == "LA" & graph_year == year(hurricane_dates$LA[1]) &
                     Date >= hurricane_dates$LA[1]) %>%
              mutate(excess = cumsum(Deaths - (s_hat + year_offset)*Population*days/365),
                     se = sqrt((s_hat + year_offset)*Population*days/365),
                     excess_se=sqrt(cumsum(se^2)),
                     lower_CI = excess-1.96*excess_se,
                     upper_CI = excess+1.96*excess_se,
                     days = cumsum(days)) %>%
              select(days, excess, excess_se, lower_CI, upper_CI, hurricane) %>%
              bind_rows(tibble(days=0, excess=0, hurricane = "LA: Katrina")) %>%
              arrange(days) %>% filter(days==katrina)
```

Percent of the deaths that occured at the day of Katrina:

```{r}
combined_monthly_res %>% 
              filter(state == "LA" & graph_year == year(hurricane_dates$LA[1]) &
                     Date >= hurricane_dates$LA[1]) %>%
              mutate(excess = cumsum(Deaths - (s_hat + year_offset)*Population*days/365),
                     se = sqrt((s_hat + year_offset)*Population*days/365),
                     excess_se=sqrt(cumsum(se^2)),
                     lower_CI = excess-1.96*excess_se,
                     upper_CI = excess+1.96*excess_se,
                     days = cumsum(days)) %>%
              select(days, excess, excess_se, lower_CI, upper_CI, hurricane) %>%
              bind_rows(tibble(days=0, excess=0, hurricane = "LA: Katrina")) %>%
              arrange(days) %>% slice(c(2,12)) %>% mutate(tmp = 100 * .$excess[1]/.$excess[2]) %>%
  select(tmp) 
```

% of deaths due to the top 4 causes of death:

```{r}
tmp <- combined_daily_res %>% # Hurricane Maria
    filter(state == "PR" & graph_year == year(hurricane_dates$PR[3]) &
             Date >= hurricane_dates$PR[3]) %>%
    mutate(excess = cumsum(Deaths - exp(offset_sample_size+offset_seasonal+offset_year)),
           days = t - min(t) + 1,
           se = sqrt(exp(offset_sample_size+offset_seasonal+offset_year)),
           excess_se=sqrt(cumsum(se^2)),
           lower_CI = excess-1.96*excess_se,
           upper_CI = excess+1.96*excess_se) %>%
    select(days, excess, excess_se, lower_CI, upper_CI, hurricane) %>%
    bind_rows(tibble(days=0, excess=0, excess_se=0, hurricane = hurricane_names$PR[3])) %>%
    arrange(days) %>% filter(days==103) %>% # Deaths until Dec 31, 2017
    .$excess

# Deaths for top 4 causes taken from table 1
deaths <- 703 + 330 + 289 + 269

# Percent
100 * deaths / tmp
```





